Glacial flows, simulated faster

### 2023-05-29 talk-iu/slides.tex
aligned better to my goals: no deep learning or Ahlkrona trick ...

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming modeling, I will ignore much of glacier physics
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (calving, crevasses)
    \item[$\circ$] solid earth deformation
    \end{itemize}

\medskip
\item {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{../images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

    \begin{itemize}
    \item[$\circ$] $\bu|_s=0$ where no ice
    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu(D\bu) = \Gamma |D\bu|^{\pp-2}$
    \end{itemize}
\end{itemize}
\end{frame}

@@

    \begin{itemize}
    \item<2>[$\circ$] but in $\infty$ dimensions (PDAE?), and subject to complementarity
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
    \begin{itemize}
    \item[$\circ$] though most researchers would agree NCP-coupled-to-Stokes is indeed the intended model \dots
    \end{itemize}
\item progress has been made on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
    \begin{itemize}
    \item[$\circ$] 1D well-posedness on flat bed (Calvo et al 2003)
    \item[$\circ$] 2D steady-state existence on general beds (Jouvet \& Bueler 2012)
    \item[$\circ$] 2D well-posedness on flat bed (Piersanti \& Temam 2022)
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item numerical modelers tend to think of the Stokes problem separately from surface evolution
    \begin{itemize}
    \item[$\circ$] \emph{time-splitting} or \emph{explicit time-stepping} is often taken for granted
    \end{itemize}

\medskip
\item ice sheet geometry evolution is addressed with minimal awareness of complementarity

\medskip
\item NCP-coupled-to-Stokes is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
    \begin{itemize}
    \item[$\circ$] because it is too slow
    \item<2>[$\circ$] \alert{can we make it fast enough to use?}
    \end{itemize}
\end{itemize}
\end{frame}

\section{time-stepping and solver scaling}

@@

\smallskip
\item<2> \emph{question:} is this really an advection equation?
\item<2>[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H$$
\item<2> NCP-coupled-to-Stokes DAE system has no characteristic curves
\end{itemize}
\end{frame}

@@

\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item the diffusion schema is literal in the lubrication approximation
    \begin{itemize}
    \item[$\circ$] more on this momentarily
    \end{itemize}
\item but the fact that ice flows downhill has \emph{time-stepping stability} consequences
    \begin{itemize}
    \item[$\circ$] regardless of your preference for the advective schema!
    \end{itemize}
\item note both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item the simplest of several shallow approximations is the ``lubrication'' approximation, the \alert{shallow ice approximation} (SIA)
\item SIA version of the NCP:
{\small
  $$s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0$$
}
the surface motion contribution ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$ has a formula:
  $${\color{FireBrick} \Phi(s)} = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)$$

@@

\medskip
\item $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
    \begin{itemize}
    \item[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
    \item[$\circ$] \emph{local} in surface and bed topography
    \item[$\circ$] existence is known for this NCP problem (Jouvet \& Bueler, 2012), when written as a \alert{variational inequality} weak form
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{time-stepping in current and future ice sheet models}

\begin{itemize}
\item current-technology large-scale models use explicit time stepping
    \begin{itemize}
    \item[$\circ$] this is embarrassing: the mathematical problem is a DAE
    \item[$\circ$] the accuracy/performance/usability consequences of the suppressed free-boundary/DAE/diffusive character are hard to sweep under the rug
    \end{itemize}
\item most researchers believe the advection schema
    \begin{itemize}
    \item[$\circ$] time step is determined by CFL using coupled solution velocity $\bU$
    \end{itemize}

\medskip
\item \alert{implicit time-stepping} is appropriate for DAE problems
\item a sequence of NCP-coupled-to-Stokes free-boundary problems must be solved at each time step
\end{itemize}
\end{frame}

@@

\section{3 approaches to better performance}

\begin{frame}{approach 1: machine learning}

\begin{itemize}
\item apply \alert{machine learning}
\item run non-scalable ice sheet models on many hypothetical/real ice sheets, and train ML \alert{emulator} on results
    \begin{itemize}
    \item[$\circ$] supervised learning of physically-based model results
    \item[$\circ$] compute map on CPUs, then learn \& evaluate map on GPUs
    \item[$\circ$] convincing demo in Jouvet et al.~(2021)
    \end{itemize}
\end{itemize}

\vspace{3mm}
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\only<1>{\includegraphics[height=0.52\textheight]{../images/tiles-jouvet.png}}
\only<2>{\includegraphics[height=0.57\textheight]{../images/cnn-jouvet.png}}
\end{center}
\end{minipage}
\end{frame}

\begin{frame}{approach 2: semi-coupled time-stepping}

\begin{itemize}
\item idea from L{\"o}fgren et al.~(2022)
    \begin{itemize}
    \item[$\circ$] earlier use in mantle/crust simulations (Kaus et al.~2010)
    \end{itemize}
\item \emph{idea.} remain explicit, but \alert{modify the Stokes problem to ``see'' the updated (extrapolated) surface}
\item that is, modify the Stokes problem at time $t^\ell$ \onslide<2>{by adding body force terms corresponding to the updated-surface icy domain}

\vspace{-3mm}
\begin{minipage}[t][30mm][t]{0.8\textwidth}
\only<1>{
{\small
\begin{align*}
\int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv &= \int_{\Lambda^\ell} \bbf\cdot \bv \\
\phantom{x} \\
\phantom{x} \\
\int_{\Lambda^\ell} q \Div \bu &= 0
\end{align*}
}
}
\only<2>{
{\small
\begin{align*}
\int_{\Lambda^\ell} 2 \nu D\bu : D\bv - \int_{\Lambda^\ell} p \Div \bv \qquad\qquad &= \int_{\Lambda^\ell} \bbf\cdot \bv + \Delta t \int_{\Gamma_s^\ell} a\, (\bbf\cdot \bv)\,dx\\
\hspace{20mm} - \Delta t \int_{\Gamma_s^\ell} (\bu \cdot \bn_{s^\ell}) (\bbf\cdot \bv)\,dx \qquad
&\phantom{=} \\
\int_{\Lambda^\ell} q \Div \bu &= 0
\end{align*}
}
}
\end{minipage}

\bigskip
\item<2> early experiments suggest $\sim 10$ times longer stable time steps
\end{itemize}

\end{frame}

\begin{frame}{approach 3: multilevel NCP-coupled-to-Stokes solvers}

\begin{itemize}
\item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs)
\item but in the non-local residual case this seems not to exist
    \begin{itemize}
    \item[$\circ$] the \alert{smoother} must reduce a residual formed from surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ (from a scalable Stokes solver)
    \end{itemize}
\item near-optimal multilevel solvers exist for simpler VI problems
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
   \begin{itemize}
   \item[$\circ$] progress is being made in all of these areas
   \end{itemize}
\item<3-> \alert{coming soon} from current research:
   \begin{enumerate}
   \item[1.] machine learning emulators (Jouvet et al.~2021)
   \item[2.] semi-coupled time stepping (L{\"o}fgren et al.~2022)
   \item[3.] scalable Stokes solvers (Isaac et al.~2015)   
   \end{enumerate}
\item<4-> scalable solvers for implicit-step NCP-coupled-to-Stokes models, which would seem to be the recommended numerical design, require \alert{multilevel solvers for non-local variational inequalities}
\end{itemize}
\end{frame}

@@

{\scriptsize
\begin{itemize}
\item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
\item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015). \emph{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics}, SIAM J.~Sci.~Comput.~37 (6), B804--B833, \href{https://doi.org/10.1137/140974407}{10.1137/140974407}
\item G.~Jouvet \& E. Bueler (2012). \emph{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation}, SIAM J.~Appl.~Math. 72 (4), 1292--1314, \href{https://doi.org/10.1137/110856654}{10.1137/110856654}
\item G.~Jouvet, G.~Cordonnier, B.~Kim, M.~L\"uthi, A.~Vieli, A.~Aschwanden (2021). \emph{Deep learning speeds up ice flow modelling by several orders of magnitude}, J.~Glaciol.~\href{https://doi.org/10.1017/jog.2021.120}{10.1017/jog.2021.120}
\item A.~L{\"o}fgren, J. Ahlkrona, \& C. Helanow (2022). \emph{Increasing stable time-step sizes of the free-surface problem arising in ice-sheet simulations}, J.~Comput.~Phys.~X 16, \href{https://doi.org/10.1016/j.jcpx.2022.100114}{10.1016/j.jcpx.2022.100114}
\end{itemize}
}
\end{frame}

>>>>

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming modeling, I will ignore parts of glacier physics in the remainder of my talk

\medskip
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (e.g.~calving)
    \item[$\circ$] solid earth deformation
    \end{itemize}

\medskip
\item {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, which includes these and other processes}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{../images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

    \begin{itemize}
    \item[$\circ$] note: $\bu|_s=0$ where no ice
    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu(D\bu) = \Gamma |D\bu|^{\pp-2}$
    \end{itemize}
\end{itemize}
\end{frame}

@@

    \begin{itemize}
    \item<2>[$\circ$] but in $\infty$ dimensions \dots and actually subject to complementarity
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
    \begin{itemize}
    \item[$\circ$] though most researchers would agree NCP-coupled-to-Stokes is indeed the intended model \dots
    \end{itemize}

\medskip
\item<2> progress has been made on well-posedness of the lubrication approximation of the basic model, the so-called \alert{shallow ice approximation}:
    \begin{itemize}
    \item[$\circ$] 1D well-posedness on flat bed (Calvo et al 2003)
    \item[$\circ$] 2D steady-state existence on general beds (Jouvet \& Bueler 2012)
    \item[$\circ$] 2D well-posedness on flat bed (Piersanti \& Temam 2022)
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item numerical glacier and ice sheet modelers tend to think of the Stokes problem separately from surface evolution
    \begin{itemize}
    \item[$\circ$] \emph{time-splitting} or \emph{explicit time-stepping} is often taken for granted
    \end{itemize}

\medskip
\item \dots and ice sheet geometry evolution is addressed with minimal awareness of complementarity

\medskip
\item the NCP-coupled-to-Stokes basic model is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
    \begin{itemize}
    \item[$\circ$] because it is too slow
    \item[$\circ$] \alert{can we make it fast enough to use?}
    \end{itemize}
\end{itemize}
\end{frame}

\section{numerics: time-stepping and solver scaling}

@@

\smallskip
\item<2> \emph{question:} is this really an advection equation?
\item<2>[] \emph{answer:} not really \dots ice flows (mostly) downhill so
  $$\bU \sim - \grad s \sim - \grad H$$
\item<2> observation: the NCP-coupled-to-Stokes system has no characteristic curves
\end{itemize}
\end{frame}

@@

\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
\begin{itemize}
\item the diffusion schema is literal in the shallow ice approximation
    \begin{itemize}
    \item[$\circ$] more on this momentarily
    \end{itemize}
\item the fact that ice flows downhill has \emph{time-stepping stability} consequences for the basic model
    \begin{itemize}
    \item[$\circ$] regardless of your above schema preference!
    \end{itemize}
\item note both forms are highly-nonlinear: \, $\bU(H,\grad s), \, D(H,\grad s)$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item as mentioned, the simplest of glacier flow approximations is the ``lubrication'' approximation: \alert{shallow ice approximation} (SIA)
\item SIA version of the NCP:
{\small
  $$s - b \ge 0, \quad \frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a \ge 0, \quad (s - b) \left(\frac{\partial s}{\partial t} + {\color{FireBrick} \Phi(s)} - a\right) = 0$$
}
the surface motion contribution ${\color{FireBrick} \Phi(s) = - \bu|_s \cdot \bn_s}$ has a formula:
  $${\color{FireBrick} \Phi(s)} = - \frac{\gamma}{\pp} (s-b)^{\pp} |\grad s|^{\pp} - \grad \cdot\left(\frac{\gamma}{\pp+1} (s-b)^{\pp+1} |\grad s|^{\pp-2} \grad s\right)$$

@@

\medskip
\item $\Phi(s)$ resolves to a \alert{doubly-nonlinear differential operator}
    \begin{itemize}
    \item[$\circ$] porous medium and $\pp$-Laplacian type simultaneously
    \item[$\circ$] \emph{local} in surface and bed topography
    \item[$\circ$] well-posedness for this problem as weak form \alert{variational inequality} (Calvo et al 2002, Jouvet \& Bueler 2012, Piersanti \& Temam 2022)
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{time-stepping in current ice sheet models}

\begin{itemize}
\item current-technology large-scale models use explicit time stepping
    \begin{itemize}
    \item[$\circ$] this is embarrassing: the mathematical problem is a DAE
    \item[$\circ$] the accuracy/performance/usability consequences of the suppressed free-boundary/DAE/diffusive character are hard to sweep under the rug
    \end{itemize}
\item yet most researchers believe the advection schema
    \begin{itemize}
    \item[$\circ$] time step is determined by CFL using coupled solution velocity $\bU$
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{time-stepping in \emph{future} ice sheet models}

\begin{itemize}
\item \alert{implicit time-stepping} is appropriate for DAE problems
\item a sequence of NCP-coupled-to-Stokes free-boundary problems must be solved at each time step
\end{itemize}
\end{frame}

@@

\section{a multilevel approach to better performance}

\begin{frame}{multilevel NCP-coupled-to-Stokes solvers}

\begin{itemize}
\item FIXME
\item direct attack on the problem seems to require a \alert{multilevel} solver for \alert{variational inequalities} (VIs)
\item but in the non-local residual case this seems not to exist
    \begin{itemize}
    \item[$\circ$] the \alert{smoother} must reduce a residual formed from surface-motion term $\Phi(s) = - \bu|_s\cdot \bn_s$ (from a scalable Stokes solver)
    \end{itemize}
\item near-optimal multilevel solvers exist for simpler VI problems
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
   \begin{itemize}
   \item[$\circ$] progress is being made in all of these areas, e.g.~ scalable Stokes solvers (Isaac et al.~2015)
   \end{itemize}
\item<3> scalable solvers for implicit-step NCP-coupled-to-Stokes models, which would seem to be the recommended numerical design, require \alert{multilevel solvers for non-local variational inequalities}
\end{itemize}
\end{frame}

@@

{\scriptsize
\begin{itemize}
\item E.~Bueler (2016). \emph{Stable finite volume element schemes for the shallow-ice approximation}, J.~Glaciol.~62 (232), 230--242, \href{https://doi.org/10.1017/jog.2015.3}{10.1017/jog.2015.3}
\item E.~Bueler (2021). \emph{Conservation laws for free-boundary fluid layers}, SIAM J.~Appl.~Math.~81 (5), 2007--2032, \href{https://doi.org/10.1137/20M135217X}{10.1137/20M135217X}
\item E.~Bueler (2022). \emph{Performance analysis of high-resolution ice-sheet simulations}, J.~Glaciol., \href{https://doi.org/10.1017/jog.2022.113}{10.1017/jog.2022.113}
\item E.~Bueler \& P.~Farrell (in preparation). \emph{A full approximation storage multilevel method for nonlinear variational inequalities}
% Calvo et al 2002
\item T.~Isaac, G.~Stadler, \& O.~Ghattas (2015). \emph{Solution of nonlinear Stokes equations discretized by high-order finite elements on nonconforming and anisotropic meshes, with application to ice sheet dynamics}, SIAM J.~Sci.~Comput.~37 (6), B804--B833, \href{https://doi.org/10.1137/140974407}{10.1137/140974407}
\item G.~Jouvet \& E.~Bueler (2012). \emph{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation}, SIAM J.~Appl.~Math.~72 (4), 1292--1314, \href{https://doi.org/10.1137/110856654}{10.1137/110856654}
\item P.~Piersanti \& R.~Temam (2022). \emph{On the dynamics of grounded shallow ice
sheets: modeling and analysis}, Adv.~Nonlinear Analysis 12, \href{https://doi.org/10.1515/anona-2022-0280}{10.1515/anona-2022-0280}
% Shapero et al ?
% Wirbel & Jarosch
\end{itemize}
}
\end{frame}
